---
title: "2023 KMeans"
author: "Kehinde Fagbamigbe"
date: "2023-01-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
#function to calculate euclidean distance
eucl.dist <- function(X, Y){
  dist <- sqrt(sum((X - Y)^2))
  return(dist) #returns euclidean distance between two points or two vectors
}

test = function(x, k){

random_center_index <- sample(1:nrow(x), size = k, replace = TRUE) #radomnly selects a row of a specified size as intial center
cluster_diff <- array(NA, dim = c(1,k)) #create a dummy array of 1 row and specified k column. This will hold the distances calculated amongst all the clusters
df <- data.frame(matrix(ncol = ncol(x), nrow = 0)) #creating a dummy dataframe of no row and same no of column as the original data

colnames(df) <- colnames(x) #renaming the column of the created dataframe with data's column name
clusters <- rep(list(df), k) #duplicating the created dataframe in the specified cluster amount


# Initialize empty clusters and old centroids
old_clusters <- rep(NA, nrow(x))
old_centroids <- matrix(NA, nrow = k, ncol = ncol(x))

while (TRUE) {
for (j in 1:nrow(x)) { #for each of the rows in the data
  for (h in 1:k) { #for each of the specified cluster
  cluster_diff[h] <- eucl.dist( x[j,] , x[random_center_index[h],] ) #calculate the euclidean distance between each observation and the randomly selected centres


  }
  
  index_of_min_centre <- which.min(cluster_diff) #find the index of the smallest euclidean distance across all the clusters
  clusters[[index_of_min_centre]] <- rbind(clusters[[index_of_min_centre]], x[j,] ) #adding to the row where the cluster belongs based on euclidean distance
}

# Check if the clusters have changed
if (identical(old_clusters, clusters)) {
  break
}

# Check for empty clusters
empty_clusters <- which(sapply(clusters, nrow) == 0)

if(length(empty_clusters) > 0) {
  # Select new random centroid for empty cluster
  random_center_index[empty_clusters] <- sample(1:nrow(x), size = length(empty_clusters), replace = TRUE)
  clusters[[empty_clusters]] <- df
}

old_clusters <- clusters
random_center_index <- sapply(clusters, function(x) x[sample(nrow(x), 1), ])


# Check for infinite loop
if (identical(old_centroids, random_center_index)) {
  break
}
old_centroids <- random_center_index


}

cluster_centres <- sapply(clusters, function(x) colMeans(x))
covariance <- sapply(clusters, function(x) cov(x))

return (list(clusters = clusters, mean = cluster_centres, cov_matrix = covariance ))
}
  
  
  

```




```{r}
source("/Users/alhajidot/Documents/BGSU/Project/EMScript.R")
```


```{r}
x <- read.table("/Users/alhajidot/Documents/BGSU/Project/gaussian.txt", quote="\"", comment.char="")
head(x)
```

```{r}
trial = test(x, 2)
trial



















```{r}
#function to calculate euclidean distance
eucl.dist <- function(X, Y){
  dist <- sqrt(sum((X - Y)^2))
  return(dist) #returns euclidean distance between two points or two vectors
  
}

test = function(x, k){
  
  random_center_index <- sample(1:nrow(x), size = k) #radomnly selects a row of a specified size as intial center
  cluster_diff <- array(NA, dim = c(1,k)) #create a dummy array of 1 row and specified k column. This will hold the distances calculated amongst all the clusters
  df <- data.frame(matrix(ncol = dim(x)[2], nrow = 0))  #creating a dummy dataframe of no row and same no of column as the original data
  
  colnames(df) <- colnames(x) #renaming the column of the created dataframe with data's column name
  clusters <- rep(list(df), k) #duplicating the created dataframe in the specified cluster amount
  
  
  # Initialize empty clusters and old centroids
  old_clusters <- rep(NA, nrow(x))
  old_centroids <- matrix(NA, nrow = k, ncol = ncol(x))
  
  while (TRUE) {
    for (j in 1:nrow(x)) { #for each of the rows in the data
      for (h in 1:k) { #for each of the specified cluster
        cluster_diff[h] <- eucl.dist( x[j,] , x[random_center_index[h],] ) #calculate the euclidean distance between each observation and  the randonly selected centres
      
      }
      index_of_min_centre <- which.min(cluster_diff) #find the index of the smallest euclidean distance across all the clusters
      clusters[[index_of_min_centre]] <- rbind(clusters[[index_of_min_centre]], x[j,] ) #adding to the row where the cluster belongs based on euclidean distance
    }
    
    # # Check if the clusters have changed
    # if (all(old_clusters == clusters)) {
    #   break
    # }
    # 
    # Check if the clusters have changed
    if (identical(old_clusters, clusters)) {
      break
    }

    
    # Check for empty clusters
    empty_clusters <- which(sapply(clusters, nrow) == 0)
    
    if(length(empty_clusters) > 0) {
      # Select new random centroid for empty cluster
      random_center_index[empty_clusters] <- sample(1:nrow(x), size = length(empty_clusters))
      clusters[[empty_clusters]] <- df
    }
    
    old_clusters <- clusters
    random_center_index <- sapply(clusters, function(x) x[[sample(nrow(x), 1)], ])


    # Check for infinite loop
    if (identical(old_centroids, random_center_index)) {
      break
    }
    old_centroids <- random_center_index
  }
  
  cluster_centres <- sapply(clusters, function(x) colMeans(x))
  covariance <- sapply(clusters, function(x) cov(x))
  
  return (list(clusters = clusters, mean = cluster_centres, cov_matrix = covariance ))
}

```



```{r}
source("/Users/alhajidot/Documents/BGSU/Project/EMScript.R")
```


```{r}
x <- read.table("/Users/alhajidot/Documents/BGSU/Project/gaussian.txt", quote="\"", comment.char="")
head(x)
```

```{r}
trial = test(x, 2)
trial
```


