---
title: "emEM Algorithm"
author: "Kehinde Fagbamigbe"
date: "2022-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
RNGkind (sample.kind = "Rounding") 
set.seed(1)
```


```{r}
x <- read.table("/Users/alhajidot/Documents/BGSU/Project/gaussian.txt", quote="\"", comment.char="")
head(x)
```
```{r}
dim(x)
```


```{r}
p = dim(x)[2]
p
```


```{r}
RNGkind (sample.kind = "Rounding") 
set.seed(1)

### Rescale the data
x.scaled <- scale(x, center = TRUE, scale = TRUE)
x.scaled.attr <- attributes(x.scaled)
x.scaled.attr
```




```{r}

cluster_centres = list()

for (k in 2:10){
    km <- kmeans(x.scaled, k, iter.max = 1)
    cluster_centres[[k]] <- c(cluster_centres, km$centers)
    #tmp.wss <- c(tmp.wss, km$tot.withinss)
    #cat("Iteration", tmp.wss, "\n")
  }
```


```{r}
covariance[[2]]
```


```{r}

cluster_centres = list()
covariance = list()
tau = list()
for (k in 2:10){
    km <- kmeans(x.scaled, k, iter.max = 1)
    cluster_centres[[k]] <- (km$centers) 
    covariance[[k]] <- (array(c(diag(p)), dim = c(p,p,k)))
    tau[[k]] = rep(1/k, k)
    #cluster_centres[[k]] <- c(cluster_centres, km$centers)
    #tmp.wss <- c(tmp.wss, km$tot.withinss)
    #cat("Iteration", tmp.wss, "\n")
    
    
    
    
    
    
  }


             # Create list
#cluster_centres 
#covariance
#tau
```





```{r}
p = dim(x)[2]
p
```

```{r}
km <- kmeans(x.scaled, k, iter.max = 1)
km$centers
class(km$centers)
dim(km$centers)
```

```{r}
km <- kmeans(x.scaled, k, iter.max = 1)
g = t(km$centers)
class(g)
dim(g)
```

```{r}
# cluster_centres = list()
# covariance = list()
# tau = list()
bic_values = list()
for (k in 2:10){
    km <- kmeans(x.scaled, k, iter.max = 1)
    mu_test <- km$centers
    g = t(km$centers)
    cov_test <- (array(c(diag(p)), dim = c(p,p,k)))
    t_test = rep(1/k, k)
    emEM = EM(x, tau = t_test,  Mu = g, covariance = cov_test, eps = 1e-4)
    bic_values[[k]] = emEM$BIC
}

bic_values
```







```{r}
```{r}
emEM = EM(x, tau = t_test,  Mu = mu_test, covariance = cov_test, eps = 1e-4)
```


```{r}
emEM$logL
```




```{r}
diag(4) * 2
```


```{r}
array(c(diag(4)), dim = c(4,4,3))
```

```{r}
array(c(diag(4)), dim = c(4,4,2))
```


```{r}
test <- kmeans(x.scaled, 2, iter.max = 1)
test$centers
```


```{r}
class(test$centers)
```


```{r}
cluster_centres[[2]]
```

```{r}
mu_test = matrix(cluster_centres[[2]], nrow = 2, ncol = 4 )
mu_test
```



```{r}
class(cluster_centres[[2]])
```


```{r}
class(mu_test)
```


```{r}
#cluster_centres
```









```{r}
#Elbow Method for finding the optimal number of clusters
set.seed(123)
# Compute and plot wss for k = 2 to k = 15.
k.max <- 15
data <- x.scaled
wss <- sapply(1:k.max, 
              function(k){kmeans(data, k, nstart=1, iter.max = 1 )$tot.withinss})
wss
```


```{r}
plot(1:k.max, wss,
     type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")
```


```{r}
k_means = kmeans(data, 6, iter.max = 1 )
k_means
```

```{r}
k_means$centers
```

```{r}
k_means$tot.withinss
```


```{r}
k_means$size
```
```{r}
k = length(k_means$size)
k
```


# Within Cluster Sum of square (wss)
# smaller wss implies within cluster homogeneity i.e closeness of data point to the centroid.

```{r}
km.list <- list()  #Empty list
wss <- NULL
centre <- NULL
seed <- NULL
cluster <- NULL
for (k in 1:3){
  tmp.wss <- NULL
  ## Running kmeans 10 times for each k to get the best result
  
  
  for (itr in 1:6){
    set.seed(itr)
    km <- kmeans(x.scaled, k)
    tmp.wss <- c(tmp.wss, km$tot.withinss)
    #cat("Iteration", tmp.wss, "\n")
  }
  
  select.seed <- which.min(tmp.wss) #which.min() returns the index of the lowest number
  #cat("Seed", select.seed, "\n")
  seed <- c(seed, select.seed)
  set.seed(select.seed) #since the index is the same as the seed we return that.
  km <- kmeans(x.scaled, k)
  cluster <- c(cluster, k)
  wss <- c(wss, km$tot.withinss)
  centre <- c(centre, km$centers)
  km.list[[k]] <- c(km.list, km)
  
}


#plot(seq(1,20,1), wss, type = "l", xlab = "Number of clusters")

```

```{r}
diag(4)
```



```{r}
centre
```


```{r}
km
```


```{r}
wss
```


```{r}
min(wss)
```

```{r}
minimum_wss_index = which(wss == min(wss))
minimum_wss_index
```


```{r}
seed
```



```{r}
seed = seed[minimum_wss_index]
seed
```

```{r}
cluster
```
```{r}
cluster = cluster[minimum_wss_index]
cluster
#cluster = 4  #remove this
```


```{r}
set.seed(seed)
km_final <- kmeans(x.scaled, cluster)
```

```{r}
km_final$centers
```



```{r}
t_test = rep(1/cluster, cluster)
t_test
```

```{r}
mu_test = t(km_final$centers)
mu_test
```

```{r}
class(mu_test)
```

```{r}
dim(mu_test)
```



```{r}
# mu_test = km_final$centers
# mu_test
```



```{r}
e1 = c(2,1,2,5)
e2 = c(9,5,8,3)
e3 = c(3,7,3,7)

cov_e1 <- matrix(cov(as.matrix(e1)), nrow = 1, ncol = 4)
cov_e1 = as.vector(cov_e1)
cv_1 = diag(cov_e1, nrow =  4, ncol = 4)
#cv_1


cov_e2 <- matrix(cov(as.matrix(e2)), nrow = 1, ncol = 4)
cov_e2 = as.vector(cov_e2)
cv_2 = diag(cov_e2, nrow =  4, ncol = 4)
#cv_2


cov_e3 <- matrix(cov(as.matrix(e3)), nrow = 1, ncol = 4)
cov_e3 = as.vector(cov_e3)
cv_3 = diag(cov_e3, nrow =  4, ncol = 4)
#cv_3


cov_test = array(c(cv_1, cv_2, cv_3), dim = c(4,4,3))
cov_test
```

```{r}
source("/Users/alhajidot/Documents/BGSU/Project/EMScript.R")
```


```{r}
#A2 <- EM(x,tau = t_test, Mu = mu_test, covariance = cov_test, eps = 1e-4)
```


```{r}
emEM = EM(x, tau = t_test,  Mu = mu_test, covariance = cov_test, eps = 1e-4)
```


```{r}
emEM$logL
```

```{r}
emEM$Mu
```

```{r}
emEM$covariance
```


```{r}
emEM$id
```

```{r}
emEM$AIC
```



```{r}
#emEM = EM(x, tau = t_test,  Mu = mu_test, covariance = cov_test, eps = 1e-4)
```




```{r}
iteration <- NULL
LogLikelihood <- NULL


for (i in 1:100) {
  set.seed(i)
  
  emEM = EM(x, tau = t_test,  Mu = mu_test, covariance = cov_test, eps = 1e-4)
  iteration <- c(iteration, i)
  LogLikelihood <- c(LogLikelihood, emEM$logL)
}
```



```{r}
LogLikelihood
```











