---
title: "Mixture Model"
author: "Kehinde Fagbamigbe"
date: "7/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#Bimodal

```{r}
x = seq(-5,12, length = 100)
x
```

```{r}
class(x)
```



```{r}
y = 0.6*dnorm(x,0,1) + 0.4*dnorm(x,5,2)
#first density of the normal distribution has a mean of zero and SD of 1

#second density of the normal distribution has a mean of 5 and SD of 2

#to obtain a bidomal type of curve, the mean of the distributions must be far apart and the variance will be small/narrow
```



```{r}
par(mar = c(4,4,1,1)+0.1) #similar to figure in python
plot(x,y, type = "l", ylab = "Density", las = 1)
```
#Trimodal
```{r}
x = seq(-5,15, length = 100)
y = 0.6*dnorm(x,0,1) + 0.4*dnorm(x,5,2) + 0.8*dnorm(x,10,1)
par(mar = c(4,4,1,1)+0.1) #similar to figure in python
plot(x,y, type = "l", ylab = "Density", las = 1)

```
# unimodal and skewed mixture of Gaussians
#Skewness

```{r}
x = seq(-5,12, length = 100)
y = 0.55*dnorm(x,0,sqrt(2)) + 0.45*dnorm(x,3,4)

#first density of the normal distribution has a mean of zero and SD of square root of 2

#second density of the normal distribution has a mean of 3 and SD of 4

#to obtain a skewed type of curve, the mean of the distributions must be close to each other and the variance will far apart


par(mar = c(4,4,1,1)+0.1) #similar to figure in python
plot(x,y, type = "l", ylab = "Density", las = 1)
```
# unimodal, symmetric and heavy tailed mixture of Gaussians
# Heavy tailed
```{r}
x = seq(-12,12, length = 100)
y = 0.4*dnorm(x,0,sqrt(2)) + 0.4*dnorm(x,0,sqrt(16)) + 0.2*dnorm(x,0,sqrt(20))
z = dnorm(x,0,sqrt( 0.4*2 + 0.4*16 + 0.2*20 )) 
par(mar = c(4,4,1,1)+0.1) #similar to figure in python
plot(x,y, type = "l", ylab = "Density", las = 1)
lines(x,z, lty = 2)
legend(2, 0.16, c("Mixture", "Gaussian"), lty = c(1,2), bty = "n", cex = 0.77)
```
## Zero Inflated Distributions (For Discrete Distribution)

# Zero-Inflated Counts
#An example of this is when you are trying to know the number of eggs in a bird's nest
```{r}

x = seq(0,15)
y = dnbinom(x,8,0.6) #dnbino is called negative binomial
#probability of successs is 0.6
#Bernoulli trials to stop after 8 successes
par(mfrow = c(2,1))
par(mar = c(4,4,1,1)+0.1) #similar to figure in python
barplot(y, names.arg = x, las = 1, xlab = "x", ylab = "Probability mass", ylim = c(0,0.22), main = "Negative Binomial", border = NA)


z = 0.2 * c(1,rep(0,length(x)-1)) + (1-0.2)*y
# 0.2 * c(1,rep(0,length(x)-1)) means giving 20% to the point mass at Zero
# (1-0.2)*y means giving 80% to the negative binomial
par(mar = c(4,4,1,1)+0.1)
barplot(z, names.arg = x, las = 1, xlab = "x", ylab = "Probability mass", ylim = c(0,0.22), main = "Zero-inflated Negative Binomial")

```

#Example of Continuous distribution: log-normal distribution, weibull distribution, exponential distribution
# Zero-Infalted Continuous Distribution
#this is the combination of the point mass at zero with a continuous distribution
#an example of this is the defect of some bulbs from the factory when you are trying to calculate the lifetime of a bulb
```{r}
x = seq(-2,15, length = 200)
y = plnorm(x,1.5,0.5) #plnorm is called density of the log normal distribution
#log normal distribution is the distribution of a number whose logarithm follows a normal distribution
#here it has a mean of 1.5 and a standard deviation of 0.5
z = 0.3 * as.numeric(x>=0) + (1-0.3)*y
par(mfrow = c(1,1))
par(mar = c(4,4,1,1)+0.1)
plot(x,y, type = "l", ylab = "Cumulative Distribution Function", xlab = "x", las = 1, lty = 2, main = "Zero-inflated log Gaussian distribution")
lines(x,z, lty = 1)
legend(4, 0.45, c("Zero infla. log Gaussian", "Log Gaussian"), lty = c(1,2), bty = "n")


```
 ## Sampling from a mixture model
 
```{r}
#Generate n observations from a mixture of two Gaussian Distributions

n = 50 #no of observation
w = c(0.6,0.4) #the weight of the two normal/gaussian distributions
mu = c(0,5) #the mean of the two normal/gaussian distributions
sigma = c(1,2) #the standard deviations of the two normal/gaussian distributions

#which component generate which of the observations

cc = sample(1:2, n, replace = T, prob = w)
x = rnorm(n, mu[cc], sigma[cc])

#Plot G(x) (the marginal density x) along the observations just sampled
xx = seq(-5,12, length = 200)
yy = w[1]*dnorm(xx, mu[1], sigma[1]) + w[2]*dnorm(xx, mu[2], sigma[2]) #Indexing in R starts from 1
par(mar = c(4,4,1,1)+0.1)
plot(xx, yy, type = "l", ylab = "Density", las = 1)
points(x, y = rep(0,n), pch = 1)
points(x, y = rep(0,n), pch = 1, col = cc)

```

# Enter code to sample 200 random numbers from a mixture of 3 Poisson distributions with means 1, 2 and 6 and weights 0.7, 0.2 # and 0.1, respectively.
 
```{r}
n = 200
w = c(0.7, 0.2, 0.1)  # Weights
lambda = c(1, 2, 6) # Means
cc = sample(1:3, n, replace=T, prob=w)
x = rpois(n, lambda[cc])
empfreq = table(factor(x, levels=seq(0, max(x))))/n
#The main difficulty here is how to properly compute the frequencies using R, rather than anything that is specific to mixture models.  Using the function table directly on the vector x will ignore values with zero counts.  Similarly, the function tabulate will ignore the x = 0 case.  First converting the vector x into a factor while ensuring that any integer between 0 and the maximum in the sample are valid factors avoids the
empfreq = table(x)/n
#Yes for some datasets, but might fail.  Use this category if, for example, the function table was used directly on x,
barplot(empfreq)
```
```{r}
x
```
 
 
# Modify the code above to sample 100 random numbers from a mixture of 4 exponential distributions with means 1, 4, 7 and 10 and weights 0.3, 0.25, 0.25 and 0.2, respectively.  Use these sample to approximate the mean and variance of the mixture.
```{r}
# # Generate n observations from a mixture of two Gaussian 
# # distributions
# n     = 100           # Size of the sample to be generated
# w     = c(0.3, 0.25, 0.25, 0.2)  # Weights
# mu    = c(1,4,7,10)      # Means
# cc    = sample(1:4, n, replace=T, prob=w)
# x = dexp(n, rate = 1/mu[cc], log = FALSE)
```
 
 
```{r}
n = 120
x = rep(0,n)
x
```
 
 
```{r}
n = 120
kk = 2
w = 0.6
mu = rep(0,kk)
mu[1] = 0
mu[2] = 5
sigma = 1
cc = sample(1:kk, n, replace = T, prob = c(w,1-w))
x = rep(0,n)
for(i in 1:n){
  x[i] = rnorm(1, mu)
}

```
 
 
```{r}
#### Example of an EM algorithm for fitting a location mixture of 2 Gaussian components
#### The algorithm is tested using simulated data

## Clear the environment and load required libraries
rm(list=ls())
set.seed(81196)    # So that results are reproducible (same simulated data every time)

## Generate data from a mixture with 2 components
KK         = 2
w.true     = 0.6  # True weights associated with the components
mu.true    = rep(0, KK)
mu.true[1] = 0   # True mean for the first component
mu.true[2] = 5   # True mean for the second component
sigma.true = 1   # True standard deviation of all components
n  = 120         # Number of observations to be generated
cc = sample(1:KK, n, replace=T, prob=c(w.true,1-w.true))
x  = rep(0, n)
for(i in 1:n){
  x[i] = rnorm(1, mu.true[cc[i]], sigma.true)  #normal distribution
}

# Plot the true density
par(mfrow=c(1,1)) #a figure frame
xx.true = seq(-8,11,length=200)
yy.true = w.true*dnorm(xx.true, mu.true[1], sigma.true) + 
          (1-w.true)*dnorm(xx.true, mu.true[2], sigma.true) 
plot(xx.true, yy.true, type="l", xlab="x", ylab="True density", lwd=2)
points(x, rep(0,n), col=cc)
```


# EM algorithm

```{r}
## Run the actual EM algorithm
## Initialize the parameters
w     = 1/2                         #Assign equal weight to each component to start with
mu    = rnorm(KK, mean(x), sd(x))   #Random cluster centers randomly spread over the support of the data
sigma = sd(x)                       #Initial standard deviation

# Plot the initial guess for the density
xx = seq(-8,11,length=200)
yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
plot(xx, yy, type="l", ylim=c(0, max(yy)), xlab="x", ylab="Initial density")
points(x, rep(0,n), col=cc)
```




```{r}
s  = 0 #just a counter of the iteration
sw = FALSE #SW = Switch. This just decides when the algorithm need to end
QQ = -Inf
QQ.out = NULL #used to understand and monitor convergence
epsilon = 10^(-5) #stopping critera for the EM Algorithm

#The EM Algorithm
##Checking convergence of the algorithm
while(!sw){
  ## E step
  v = array(0, dim=c(n,KK))
  v[,1] = log(w) + dnorm(x, mu[1], sigma, log=TRUE)    #Compute the log of the weights
  v[,2] = log(1-w) + dnorm(x, mu[2], sigma, log=TRUE)  #Compute the log of the weights
  
  for(i in 1:n){
    v[i,] = exp(v[i,] - max(v[i,]))/sum(exp(v[i,] - max(v[i,])))  #Go from logs to actual weights in a numerically stable manner
  }
  
## M step
  # Weights
  w = mean(v[,1])
  mu = rep(0, KK)
  for(k in 1:KK){
    for(i in 1:n){
      mu[k]    = mu[k] + v[i,k]*x[i]
    }
    mu[k] = mu[k]/sum(v[,k])
  }
  # Standard deviations
  sigma = 0
  for(i in 1:n){
    for(k in 1:KK){
      sigma = sigma + v[i,k]*(x[i] - mu[k])^2
    }
  }
  sigma = sqrt(sigma/sum(v))

##Check convergence
  QQn = 0
  for(i in 1:n){
    QQn = QQn + v[i,1]*(log(w) + dnorm(x[i], mu[1], sigma, log=TRUE)) +
                v[i,2]*(log(1-w) + dnorm(x[i], mu[2], sigma, log=TRUE))
  }
  if(abs(QQn-QQ)/abs(QQn)<epsilon){
    sw=TRUE
  }
  QQ = QQn
  QQ.out = c(QQ.out, QQ)
  s = s + 1
  print(paste(s, QQn))

#Plot current estimate over data
  layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
  par(mar=c(3.1,4.1,0.5,0.5))
  plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q", lwd=2)
  
  par(mar=c(5,4,1.5,0.5))
  xx = seq(-8,11,length=200)
  yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
  plot(xx, yy, type="l", ylim=c(0, max(c(yy,yy.true))), main=paste("s =",s,"   Q =", round(QQ.out[s],4)), lwd=2, col="red", lty=2, xlab="x", ylab="Density")
  lines(xx.true, yy.true, lwd=2)
  points(x, rep(0,n), col=cc)
  legend(6,0.22,c("Truth","Estimate"),col=c("black","red"), lty=c(1,2))
}
```




```{r}
#Plot final estimate over data
layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
par(mar=c(3.1,4.1,0.5,0.5)) #figure for plot
plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q", lwd=2)

par(mar=c(5,4,1.5,0.5))
xx = seq(-8,11,length=200)
yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
plot(xx, yy, type="l", ylim=c(0, max(c(yy,yy.true))), main=paste("s =",s,"   Q =", round(QQ.out[s],4)), lwd=2, col="red", lty=2, xlab="x", ylab="Density")
lines(xx.true, yy.true, lwd=2)
points(x, rep(0,n), col=cc)
legend(6,0.22,c("Truth","Estimate"),col=c("black","red"), lty=c(1,2), bty="n")


```
 
```{r}
#### Example of an EM algorithm for fitting a mixtures of K p-variate Gaussian components
#### The algorithm is tested using simulated data

## Clear the environment and load required libraries
rm(list=ls())
library(mvtnorm)    # Multivariate normals are not default in R
library(ellipse)    # Required for plotting
set.seed(63252)     # For reproducibility

## Generate data from a mixture with 3 components
KK      = 3
p       = 2
w.true = c(0.5,0.3,0.2)  # True weights associated with the components
mu.true     = array(0, dim=c(KK,p))
mu.true[1,] = c(0,0)   #True mean for the first component
mu.true[2,] = c(5,5)   #True mean for the second component
mu.true[3,] = c(-3,7)   #True mean for the third component
Sigma.true      = array(0, dim=c(KK,p,p))
Sigma.true[1,,] = matrix(c(1,0,0,1),p,p)   #True variance for the first component
Sigma.true[2,,] = matrix(c(2,0.9,0.9,1),p,p)   #True variance for the second component
Sigma.true[3,,] = matrix(c(1,-0.9,-0.9,4),p,p)   #True variance for the third component
n  = 120
cc = sample(1:3, n, replace=T, prob=w.true)
x  = array(0, dim=c(n,p))
for(i in 1:n){
  x[i,] = rmvnorm(1, mu.true[cc[i],], Sigma.true[cc[i],,])
}
```


```{r}
par(mfrow=c(1,1))
plot(x[,1], x[,2], col=cc, type="n", xlab=expression(x[1]), ylab=expression(x[2]))
text(x[,1], x[,2], seq(1,n), col=cc, cex=0.6)


for(k in 1:KK){
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.50), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.82), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.95), col="grey", lty=2, lwd=2)
  
}
title(main="Data + True Components")
```


```{r}
### Run the EM algorithm
## Initialize the parameters
w   = rep(1,KK)/KK  #Assign equal weight to each component to start with
mu  = rmvnorm(KK, apply(x,2,mean), var(x))   #RandomCluster centers randomly spread over the support of the data
Sigma      = array(0, dim=c(KK,p,p))  #Initial variances are assumed to be the same
Sigma[1,,] = var(x)/KK  
Sigma[2,,] = var(x)/KK
Sigma[3,,] = var(x)/KK

par(mfrow=c(1,1))
plot(x[,1], x[,2], col=cc, xlab=expression(x[1]), ylab=expression(x[2]))
for(k in 1:KK){
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.50), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.82), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.95), col="grey", lty=2, lwd=2)
}
```


```{r}
s       = 0
sw      = FALSE
QQ      = -Inf
QQ.out  = NULL
epsilon = 10^(-5)

while(!sw){
  ## E step
  v = array(0, dim=c(n,KK))
  for(k in 1:KK){
    v[,k] = log(w[k]) + dmvnorm(x, mu[k,], Sigma[k,,],log=TRUE)  #Compute the log of the weights
  }
  for(i in 1:n){
    v[i,] = exp(v[i,] - max(v[i,]))/sum(exp(v[i,] - max(v[i,])))  #Go from logs to actual weights in a numerically stable manner
  }
  
  ## M step
  w     = apply(v,2,mean)
  mu    = array(0, dim=c(KK, p))
  for(k in 1:KK){
    for(i in 1:n){
      mu[k,]    = mu[k,] + v[i,k]*x[i,]
    }
    mu[k,] = mu[k,]/sum(v[,k])
  }
  Sigma = array(0, dim=c(KK, p, p))
  for(k in 1:KK){
    for(i in 1:n){
      Sigma[k,,] = Sigma[k,,] + v[i,k]*(x[i,] - mu[k,])%*%t(x[i,] - mu[k,])
    }
    Sigma[k,,] = Sigma[k,,]/sum(v[,k])
  }
  
  ##Check convergence
  QQn = 0
  for(i in 1:n){
    for(k in 1:KK){
      QQn = QQn + v[i,k]*(log(w[k]) + dmvnorm(x[i,],mu[k,],Sigma[k,,],log=TRUE))
    }
  }
  if(abs(QQn-QQ)/abs(QQn)<epsilon){
    sw=TRUE
  }
  QQ = QQn
  QQ.out = c(QQ.out, QQ)
  s = s + 1
  print(paste(s, QQn))
  
  #Plot current components over data
  layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
  par(mar=c(3.1,4.1,0.5,0.5))
  plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q")
  
  par(mar=c(5,4,1,0.5))
  plot(x[,1], x[,2], col=cc, main=paste("s =",s,"   Q =", round(QQ.out[s],4)), 
       xlab=expression(x[1]), ylab=expression(x[2]), lwd=2)
  for(k in 1:KK){
    lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.50), col="grey", lty=2, lwd=2)
    lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.82), col="grey", lty=2, lwd=2)
    lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.95), col="grey", lty=2, lwd=2)
  }
}

#Plot current components over data
layout(matrix(c(1,2),2,1), widths=c(1,1), heights=c(1.3,3))
par(mar=c(3.1,4.1,0.5,0.5))
plot(QQ.out[1:s],type="l", xlim=c(1,max(10,s)), las=1, ylab="Q", lwd=2)

par(mar=c(5,4,1,0.5))
plot(x[,1], x[,2], col=cc, main=paste("s =",s,"   Q =", round(QQ.out[s],4)), xlab=expression(x[1]), ylab=expression(x[2]))
for(k in 1:KK){
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.50), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.82), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.95), col="grey", lty=2, lwd=2)
}





```
 
 
```{r}
mu
```
 
 
 
```{r}
mu.true
```
 
 
 
```{r}
v[58,]
```




#### Example of an MCMC algorithm for fitting a location mixture of 2 Gaussian components
#### The algorithm is tested using simulated data

```{r}
## Clear the environment and load required libraries
rm(list=ls())
library(MCMCpack)
set.seed(81196)  # So that results are reproducible

## Generate data from a mixture with 2 components
KK         = 2
w.true     = 0.6  # True weights associated with the components
mu.true    = rep(0, KK)
mu.true[1] = 0   # True mean for the first component
mu.true[2] = 5   # True mean for the second component
sigma.true = 1   # True standard deviation of all components
n          = 120         # Number of observations to be generated
cc.true = sample(1:KK, n, replace=T, prob=c(w.true,1-w.true))
x  = rep(0, n)
for(i in 1:n){
  x[i] = rnorm(1, mu.true[cc.true[i]], sigma.true)
}

# Plot the true density
par(mfrow=c(1,1))
xx.true = seq(-8,11,length=200)
yy.true = w.true*dnorm(xx.true, mu.true[1], sigma.true) + 
  (1-w.true)*dnorm(xx.true, mu.true[2], sigma.true) 
plot(xx.true, yy.true, type="l", xlab="x", ylab="True density", lwd=2)
points(x, rep(0,n), col=cc.true)
```


#Initial guess of the data distribution

```{r}
## Initialize the parameters
w     = 1/2                         #Assign equal weight to each component to start with
mu    = rnorm(KK, mean(x), sd(x))   #Random cluster centers randomly spread over the support of the data
sigma = sd(x)                       #Initial standard deviation

# Plot the initial guess for the density
xx = seq(-8,11,length=200)
yy = w*dnorm(xx, mu[1], sigma) + (1-w)*dnorm(xx, mu[2], sigma)
plot(xx, yy, type="l", ylim=c(0, max(yy)), xlab="x", 
     ylab="Initial density", lwd=2)
points(x, rep(0,n), col=cc.true)
```


```{r}
## The actual MCMC algorithm starts here
# Priors
aa  = rep(1,KK)  # Uniform prior on w
eta = 0          # Mean 0 for the prior on mu_k
tau = 5          # Standard deviation 5 on the prior for mu_l
dd  = 2          # no of degree of freedom for the prior on the variance 
qq  = 1          #the other parameter for the prior on the variance is going to be 1

# Number of iterations of the sampler
rrr   = 6000 #no of iteration of the MCMC
burn  = 1000


# Storing the samples
cc.out    = array(0, dim=c(rrr, n))   #Contains the indicators
w.out     = rep(0, rrr)               # stores the samples of the weight
mu.out    = array(0, dim=c(rrr, KK))  # stores the samples of the means of the components
sigma.out = rep(0, rrr)               # stores the samples of the standard deviation of the components
logpost   = rep(0, rrr)               # contains the logarithm of the posterior

# MCMC iterations
for(s in 1:rrr){
  # Sample the indicators
  cc = rep(0,n)
  for(i in 1:n){
    v = rep(0,KK)
    v[1] = log(w) + dnorm(x[i], mu[1], sigma, log=TRUE)  #Compute the log of the weights
    v[2] = log(1-w) + dnorm(x[i], mu[2], sigma, log=TRUE)  #Compute the log of the weights
    v = exp(v - max(v))/sum(exp(v - max(v)))
    cc[i] = sample(1:KK, 1, replace=TRUE, prob=v)   #sampling the mixture components
  }
  
  # Sample the weights
  w = rbeta(1, aa[1] + sum(cc==1), aa[2] + sum(cc==2))

  # Sample the means
  for(k in 1:KK){
    nk    = sum(cc==k)
    xsumk = sum(x[cc==k])
    tau2.hat = 1/(nk/sigma^2 + 1/tau^2)
    mu.hat  = tau2.hat*(xsumk/sigma^2 + eta/tau^2)
    mu[k]   = rnorm(1, mu.hat, sqrt(tau2.hat))
  }

  # Sample the variances
  dd.star = dd + n/2
  qq.star = qq + sum((x - mu[cc])^2)/2
  sigma = sqrt(rinvgamma(1, dd.star, qq.star))

  # Store samples
  #Restoring the values we just sampled
  cc.out[s,]   = cc
  w.out[s]     = w
  mu.out[s,]   = mu
  sigma.out[s] = sigma
  
  #computing the normalized posterior for the model
  for(i in 1:n){
    if(cc[i]==1){
      logpost[s] = logpost[s] + log(w) + dnorm(x[i], mu[1], sigma, log=TRUE)
    }else{
      logpost[s] = logpost[s] + log(1-w) + dnorm(x[i], mu[2], sigma, log=TRUE)
    }
  }
  logpost[s] = logpost[s] + dbeta(w, aa[1], aa[2],log = T)
  for(k in 1:KK){
    logpost[s] = logpost[s] + dnorm(mu[k], eta, tau, log = T)
  }
  logpost[s] = logpost[s] + log(dinvgamma(sigma^2, dd, 1/qq))
  
  #print out every 500 iteration-what the current iteration is
  if(s/500==floor(s/500)){
    print(paste("s =",s))
  }
}
```


```{r}
#to verify convergence
#plot the log posterior with iteration number
## Plot the logposterior distribution for various samples
par(mfrow=c(1,1))
par(mar=c(4,4,1,1)+0.1)
plot(logpost, type="l", xlab="Iterations", ylab="Log posterior")
```


```{r}
## ---- mcmunivariate4
#plot of the posterior distribution
xx = seq(-8,11,length=200)
density.posterior = array(0, dim=c(rrr-burn,length(xx)))
for(s in 1:(rrr-burn)){
  density.posterior[s,] = density.posterior[s,] + w.out[s+burn]*dnorm(xx,mu.out[s+burn,1],sigma.out[s+burn]) +
                                                  (1-w.out[s+burn])*dnorm(xx,mu.out[s+burn,2],sigma.out[s+burn])
}
density.posterior.m = apply(density.posterior , 2, mean)
density.posterior.lq = apply(density.posterior, 2, quantile, 0.025)
density.posterior.uq = apply(density.posterior, 2, quantile, 0.975)
par(mfrow=c(1,1))
par(mar=c(4,4,1,1)+0.1)
plot(xx, density.posterior.m, type="n",ylim=c(0,max(density.posterior.uq)), xlab="x", ylab="Density")
polygon(c(xx,rev(xx)), c(density.posterior.lq, rev(density.posterior.uq)), col="grey", border="grey")
lines(xx, density.posterior.m, lwd=2)
points(x, rep(0,n), col=cc.true)
```
# the black line corresponds to the posterior mean of the density of data
# the gray area corresponds to critical intervals- point wise credible intervals for that density




#### Example of an MCMC algorithm for fitting a mixtures of K p-variate Gaussian components
#### The algorithm is tested using simulated data



```{r}
## Clear the environment and load required libraries
rm(list=ls())
library(mvtnorm)
library(ellipse)
library(MCMCpack)

## Generate data from a mixture with 3 components
#three normals - bivariate normals
KK      = 3
p       = 2
w.true = c(0.5,0.3,0.2)  # True weights associated with the components
mu.true     = array(0, dim=c(KK,p))
mu.true[1,] = c(0,0)   #True mean for the first component
mu.true[2,] = c(5,5)   #True mean for the second component
mu.true[3,] = c(-3,7)   #True mean for the third component
Sigma.true      = array(0, dim=c(KK,p,p))
Sigma.true[1,,] = matrix(c(1,0,0,1),p,p)   #True variance for the first component
Sigma.true[2,,] = matrix(c(2,0.9,0.9,1),p,p)   #True variance for the second component
Sigma.true[3,,] = matrix(c(1,-0.9,-0.9,4),p,p)   #True variance for the third component
set.seed(63252)    #Keep seed the same so that we can reproduce results
n  = 120
cc.true = sample(1:3, n, replace=T, prob=w.true)
x  = array(0, dim=c(n,p))
for(i in 1:n){
  x[i,] = rmvnorm(1, mu.true[cc.true[i],], Sigma.true[cc.true[i],,])
}

par(mfrow=c(1,1))
plot(x[,1], x[,2], col=cc.true, xlab=expression(x[1]), ylab=expression(x[2]), type="n")
text(x[,1], x[,2], seq(1,n), col=cc.true, cex=0.6)
for(k in 1:KK){
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.50), col="grey", lty=2)
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.82), col="grey", lty=2)
  lines(ellipse(x=Sigma.true[k,,], centre=mu.true[k,], level=0.95), col="grey", lty=2)
}
title(main="Data + True Components")
```

# Initial Guess
```{r}
## Initialize the parameters
w          = rep(1,KK)/KK  #Assign equal weight to each component to start with
mu         = rmvnorm(KK, apply(x,2,mean), var(x))   #RandomCluster centers randomly spread over the support of the data (rmvnorm means mulitvariate normal distribution)
Sigma      = array(0, dim=c(KK,p,p))  #Initial variances are assumed to be the same
Sigma[1,,] = var(x)/KK  
Sigma[2,,] = var(x)/KK
Sigma[3,,] = var(x)/KK
cc         = sample(1:KK, n, replace=TRUE, prob=w)

par(mfrow=c(1,1))
plot(x[,1], x[,2], col=cc.true, xlab=expression(x[1]), ylab=expression(x[2]))
for(k in 1:KK){
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.50), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.82), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.95), col="grey", lty=2, lwd=2)
}
title(main="Initial estimate + Observations")
```


```{r}
# Priors
aa = rep(1, KK)
dd = apply(x,2,mean)
DD = 10*var(x)
nu = p
SS = var(x)/3

# Number of iteration of the sampler
rrr = 1000

# Storing the samples
cc.out    = array(0, dim=c(rrr, n))
w.out     = array(0, dim=c(rrr, KK))
mu.out    = array(0, dim=c(rrr, KK, p))
Sigma.out = array(0, dim=c(rrr, KK, p, p))
logpost   = rep(0, rrr)

for(s in 1:rrr){
  # Sample the indicators
  for(i in 1:n){
    v = rep(0,KK)
    for(k in 1:KK){
      v[k] = log(w[k]) + mvtnorm::dmvnorm(x[i,], mu[k,], Sigma[k,,], log=TRUE)  #Compute the log of the weights
    }
    v = exp(v - max(v))/sum(exp(v - max(v)))
    cc[i] = sample(1:KK, 1, replace=TRUE, prob=v)
  }
  
  # Sample the weights
  w = as.vector(rdirichlet(1, aa + tabulate(cc)))
  
  # Sample the means
  DD.st = matrix(0, nrow=p, ncol=p)
  for(k in 1:KK){
    mk    = sum(cc==k)
    xsumk = apply(x[cc==k,], 2, sum)
    DD.st = solve(mk*solve(Sigma[k,,]) + solve(DD))
    dd.st = DD.st%*%(solve(Sigma[k,,])%*%xsumk + solve(DD)%*%dd)
    mu[k,] = as.vector(rmvnorm(1,dd.st,DD.st))
  }
  
  # Sample the variances
  xcensumk = array(0, dim=c(KK,p,p))
  for(i in 1:n){
    xcensumk[cc[i],,] = xcensumk[cc[i],,] + (x[i,] - mu[cc[i],])%*%t(x[i,] - mu[cc[i],])
  }
  for(k in 1:KK){
    Sigma[k,,] = riwish(nu + sum(cc==k), SS + xcensumk[k,,])
  }
  
  # Store samples
  cc.out[s,]      = cc
  w.out[s,]       = w
  mu.out[s,,]     = mu
  Sigma.out[s,,,] = Sigma
  for(i in 1:n){
    logpost[s] = logpost[s] + log(w[cc[i]]) + mvtnorm::dmvnorm(x[i,], mu[cc[i],], Sigma[cc[i],,], log=TRUE)
  }
  logpost[s] = logpost[s] + ddirichlet(w, aa)
  for(k in 1:KK){
    logpost[s] = logpost[s] + mvtnorm::dmvnorm(mu[k,], dd, DD, log=TRUE)
    logpost[s] = logpost[s] + log(diwish(Sigma[k,,], nu, SS))
  }
  
  if(s/250==floor(s/250)){
    print(paste("s = ", s))
  }
  
}

## Plot the logposterior distribution for various samples
par(mfrow=c(1,1))
par(mar=c(4,4,1,1)+0.1)
plot(logpost, type="l", xlab="Iterations", ylab="Log posterior")

## Plot the density estimate for the last iteration of the MCMC
par(mfrow=c(1,1))
par(mar=c(4,4,2,1)+0.1)
plot(x[,1], x[,2], col=cc.true, main=paste("s =",s,"   logpost =", round(logpost[s],4)), xlab=expression(x[1]), ylab=expression(x[2]))
for(k in 1:KK){
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.50), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.82), col="grey", lty=2, lwd=2)
  lines(ellipse(x=Sigma[k,,], centre=mu[k,], level=0.95), col="grey", lty=2, lwd=2)
}
```
# Application of Mixture Model

1. Density Estimation
```{r}
## Using mixture models for density estimation in the galaxies dataset
## Compare kernel density estimation, and estimates from mixtures of KK=6 
## components obtained using both frequentist and Bayesian procedures
```

#aim
# we want to know if the distribution is multimodal
# how many modes are presnt
# provide the density estimator for this different velocities


# Load the galaxy dataset
```{r}
rm(list=ls())

### Loading data and setting up global variables
library(MASS)
library(MCMCpack)
data(galaxies)
KK = 6          # Based on the description of the dataset
x  = galaxies
n  = length(x)
set.seed(781209)
```


```{r}
x
```


#using the EM algorithm to get the maximum likelihood estimator of the parameters of this particular dataset
# once the maximum likelihood estimators are plugged into the definition of the mixture model, we can obtain what we call the
# Maximum Likelihood density estimate based on the location of the mixture of models

```{r}
### First, compute the "Maximum Likelihood" density estimate associated with a location mixture of 6 Gaussian distributions using the EM algorithm
#Initial Guess
## Initialize the parameters
w     = rep(1,KK)/KK
mu    = rnorm(KK, mean(x), sd(x))
sigma = sd(x)/KK


#initialization of parameters that control the convergence of the algorithm
epsilon = 0.000001
s       = 0
sw      = FALSE
KL      = -Inf
KL.out  = NULL

while(!sw){
  ## E step
  v = array(0, dim=c(n,KK))
  for(k in 1:KK){
    v[,k] = log(w[k]) + dnorm(x, mu[k], sigma,log=TRUE)  
  }
  for(i in 1:n){
    v[i,] = exp(v[i,] - max(v[i,]))/sum(exp(v[i,] - max(v[i,])))
  }
  
  ## M step
  # Weights
  w = apply(v,2,mean)
  mu = rep(0, KK)
  for(k in 1:KK){
    for(i in 1:n){
      mu[k]    = mu[k] + v[i,k]*x[i]
    }
    mu[k] = mu[k]/sum(v[,k])
  }
  # Standard deviations
  sigma = 0
  for(i in 1:n){
    for(k in 1:KK){
      sigma = sigma + v[i,k]*(x[i] - mu[k])^2
    }
  }
  sigma = sqrt(sigma/sum(v))
  
  ##Check convergence
  KLn = 0
  for(i in 1:n){
    for(k in 1:KK){
      KLn = KLn + v[i,k]*(log(w[k]) + dnorm(x[i], mu[k], sigma, log=TRUE))
    }
  }
  if(abs(KLn-KL)/abs(KLn)<epsilon){
    sw=TRUE
  }
  KL = KLn
  KL.out = c(KL.out, KL)
  s = s + 1
  print(paste(s, KLn))
}
xx  = seq(5000,37000,length=300)
nxx = length(xx)
density.EM = rep(0, nxx)
for(s in 1:nxx){
  for(k in 1:KK){
    density.EM[s] = density.EM[s] + w[k]*dnorm(xx[s], mu[k], sigma)
  }
}
```


```{r}
### Get a "Bayesian" kernel density estimator based on the same location mixture of 6 normals
## Priors set up using an "empirical Bayes" approach
aa  = rep(1,KK)  
eta = mean(x)    
tau = sqrt(var(x))
dd  = 2
qq  = var(x)/KK

## Initialize the parameters
w     = rep(1,KK)/KK
mu    = rnorm(KK, mean(x), sd(x))
sigma = sd(x)/KK
cc    = sample(1:KK, n, replace=T, prob=w)

## Number of iterations of the sampler
rrr   = 12000
burn  = 2000

## Storing the samples
cc.out    = array(0, dim=c(rrr, n))
w.out     = array(0, dim=c(rrr, KK))
mu.out    = array(0, dim=c(rrr, KK))
sigma.out = rep(0, rrr)
logpost   = rep(0, rrr)


for(s in 1:rrr){
  # Sample the indicators
  for(i in 1:n){
    v = rep(0,KK)
    for(k in 1:KK){
      v[k] = log(w[k]) + dnorm(x[i], mu[k], sigma, log=TRUE)  #Compute the log of the weights
    }
    v = exp(v - max(v))/sum(exp(v - max(v)))
    cc[i] = sample(1:KK, 1, replace=TRUE, prob=v)
  }
  
  # Sample the weights
  w = as.vector(rdirichlet(1, aa + tabulate(cc, nbins=KK)))
  
  # Sample the means
  for(k in 1:KK){
    nk    = sum(cc==k)
    xsumk = sum(x[cc==k])
    tau2.hat = 1/(nk/sigma^2 + 1/tau^2)
    mu.hat  = tau2.hat*(xsumk/sigma^2 + eta/tau^2)
    mu[k]   = rnorm(1, mu.hat, sqrt(tau2.hat))
  }
  
  # Sample the variances
  dd.star = dd + n/2
  qq.star = qq + sum((x - mu[cc])^2)/2
  sigma = sqrt(1/rgamma(1, dd.star, qq.star))
  
  # Store samples
  cc.out[s,]   = cc
  w.out[s,]    = w
  mu.out[s,]   = mu
  sigma.out[s] = sigma
  for(i in 1:n){
    logpost[s] = logpost[s] + log(w[cc[i]]) + dnorm(x[i], mu[cc[i]], sigma, log=TRUE)
  }
  logpost[s] = logpost[s] + log(ddirichlet(w, aa))
  for(k in 1:KK){
    logpost[s] = logpost[s] + dnorm(mu[k], eta, tau, log=TRUE)
  }
  logpost[s] = logpost[s] + dgamma(1/sigma^2, dd, qq, log=TRUE) - 4*log(sigma)
  if(s/500==floor(s/500)){
    print(paste("s =",s))
  }
}

## Compute the samples of the density over a dense grid
density.mcmc = array(0, dim=c(rrr-burn,length(xx)))
for(s in 1:(rrr-burn)){
    for(k in 1:KK){
        density.mcmc[s,] = density.mcmc[s,] + w.out[s+burn,k]*dnorm(xx,mu.out[s+burn,k],sigma.out[s+burn])
    }
}
density.mcmc.m = apply(density.mcmc , 2, mean)
```


```{r}
## Plot Bayesian estimate with pointwise credible bands along with kernel density estimate and frequentist point estimate
colscale = c("black", "blue", "red")
yy = density(x)
density.mcmc.lq = apply(density.mcmc, 2, quantile, 0.025)
density.mcmc.uq = apply(density.mcmc, 2, quantile, 0.975)
plot(xx, density.mcmc.m, type="n",ylim=c(0,max(density.mcmc.uq)),xlab="Velocity", ylab="Density")
polygon(c(xx,rev(xx)), c(density.mcmc.lq, rev(density.mcmc.uq)), col="grey", border="grey")
lines(xx, density.mcmc.m, col=colscale[1], lwd=2)
lines(xx, density.EM, col=colscale[2], lty=2, lwd=2)
lines(yy, col=colscale[3], lty=3, lwd=2)
points(x, rep(0,n))
legend(27000, 0.00017, c("KDE","EM","MCMC"), col=colscale[c(3,2,1)], lty=c(3,2,1), lwd=2, bty="n")

```

# Clustering
```{r}
## Using mixture models for clustering in the iris dataset
## Compare k-means clustering and a location and scale mixture model with K normals

### Loading data and setting up global variables
rm(list=ls())
library(mclust)
library(mvtnorm)

### Defining a custom function to create pair plots
### This is an alternative to the R function pairs() that allows for 
### more flexibility. In particular, it allows us to use text to label 
### the points
pairs2 = function(x, col="black", pch=16, labels=NULL, names = colnames(x)){
  n = dim(x)[1]
  p = dim(x)[2]
  par(mfrow=c(p,p))
  for(k in 1:p){
    for(l in 1:p){
      if(k!=l){
        par(mar=c(3,3,1,1)+0.1)
        plot(x[,k], x[,l], type="n", xlab="", ylab="")
        if(is.null(labels)){
          points(x[,k], x[,l], pch=pch, col=col)
        }else{
          text(x[,k], x[,l], labels=labels, col=col)
        }
      }else{
        plot(seq(0,5), seq(0,5), type="n", xlab="", ylab="", axes=FALSE)
        text(2.5,2.5,names[k], cex=1.2)
      }
    }
  }
}

## Setup data
data(iris)
x       = as.matrix(iris[,-5])
n       = dim(x)[1]
p       = dim(x)[2]       # Number of features
KK      = 3
epsilon = 0.0000001
par(mfrow=c(1,1))
par(mar=c(4,4,1,1))
colscale = c("black","blue","red")
shortnam  = c("s","c","g")
pairs2(x, col=colscale[iris[,5]], labels=shortnam[as.numeric(iris[,5])])
```




```{r}
# Initialize the parameters of the algorithm
set.seed(63252)
numruns = 15
v.sum   = array(0, dim=c(numruns, n, KK))
QQ.sum  = rep(0, numruns)

for(ss in 1:numruns){
  w   = rep(1,KK)/KK  #Assign equal weight to each component to start with
  mu  = rmvnorm(KK, apply(x,2,mean), 3*var(x))   #Cluster centers randomly spread over the support of the data
  Sigma      = array(0, dim=c(KK,p,p))  #Initial variances are assumed to be the same
  Sigma[1,,] = var(x)
  Sigma[2,,] = var(x)
  Sigma[3,,] = var(x)
  
  sw     = FALSE
  QQ     = -Inf
  QQ.out = NULL
  s      = 0
  
  while(!sw){
    ## E step
    v = array(0, dim=c(n,KK))
    for(k in 1:KK){  #Compute the log of the weights
      v[,k] = log(w[k]) + mvtnorm::dmvnorm(x, mu[k,], Sigma[k,,], log=TRUE) 
    }
    for(i in 1:n){
      v[i,] = exp(v[i,] - max(v[i,]))/sum(exp(v[i,] - max(v[i,])))  #Go from logs to actual weights in a numerically stable manner
    }
    
    ## M step
    w = apply(v,2,mean)
    mu = matrix(0, nrow=KK, ncol=p)
    for(k in 1:KK){
      for(i in 1:n){
        mu[k,]    = mu[k,] + v[i,k]*x[i,]
      }
      mu[k,] = mu[k,]/sum(v[,k])
    }
    Sigma = array(0,dim=c(KK, p, p))
    for(k in 1:KK){
      for(i in 1:n){
        Sigma[k,,] = Sigma[k,,] + v[i,k]*(x[i,] - mu[k,])%*%t(x[i,] - mu[k,])
      }
      Sigma[k,,] = Sigma[k,,]/sum(v[,k])
    }
    
    ##Check convergence
    QQn = 0
    for(i in 1:n){
      for(k in 1:KK){
        QQn = QQn + v[i,k]*(log(w[k]) + mvtnorm::dmvnorm(x[i,],mu[k,],Sigma[k,,],log=TRUE))
      }
    }
    if(abs(QQn-QQ)/abs(QQn)<epsilon){
      sw=TRUE
    }
    QQ = QQn
    QQ.out = c(QQ.out, QQ)
    s = s + 1
  }
  
  v.sum[ss,,] = v
  QQ.sum[ss]  = QQ.out[s]
  print(paste("ss =", ss))
}
```





```{r}
## Cluster reconstruction under the mixture model
cc = apply(v.sum[which.max(QQ.sum),,], 1 ,which.max)
colscale = c("black","blue","red")
pairs2(x, col=colscale[cc], labels=cc)
```

# Rand Index
```{r}
ARImle = adjustedRandIndex(cc, as.numeric(iris[,5]))  # Higher values indicate larger agreement
ARImle
```


```{r}
## Cluster reconstruction under the K-means algorithm
irisCluster <- kmeans(x, 3, nstart = numruns)
colscale = c("black","blue","red")
pairs2(x, col=colscale[irisCluster$cluster], labels=irisCluster$cluster)
```


```{r}
ARIkmeans = adjustedRandIndex(irisCluster$cluster, as.numeric(iris[,5]))
ARIkmeans
```

