---
title: "Preliminary Analysis"
author: "Kehinde Fagbamigbe"
date: "2023-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
data <- read.csv("~/Documents/BGSU/Project/hd2019.csv")
head(data)
```
```{r}
tail(data)
```


```{r}
colnames(data)
```

UNITID where data is found for merged schools.
INSTNM - Institution (entity) name
IALIAS - Institution name alias
- ADDR - Institution (entity) street address or post office box
- CITY
- STABBR - State Abbreviation
- ZIP -
- FIPS -	Frequency	Percentage State Code

Value	Label	Frequency	Percentage
1	Alabama		
2	Alaska		
4	Arizona		
5	Arkansas		
6	California		
8	Colorado		
9	Connecticut		
10	Delaware		
11	District of Columbia		
12	Florida		
13	Georgia		
15	Hawaii		
16	Idaho		
17	Illinois		
18	Indiana		
19	Iowa		
20	Kansas		
21	Kentucky		
22	Louisiana		
23	Maine		
24	Maryland		
25	Massachusetts		
26	Michigan		
27	Minnesota		
28	Mississippi		
29	Missouri		
30	Montana		
31	Nebraska		
32	Nevada		
33	New Hampshire		
34	New Jersey		
35	New Mexico		
36	New York		
37	North Carolina		
38	North Dakota		
39	Ohio		
40	Oklahoma		
41	Oregon		
42	Pennsylvania		
44	Rhode Island		
45	South Carolina		
46	South Dakota		
47	Tennessee		
48	Texas		
49	Utah		
50	Vermont		
51	Virginia		
53	Washington		
54	West Virginia		
55	Wisconsin		
56	Wyoming		
60	American Samoa		
64	Federated States of Micronesia		
66	Guam		
68	Marshall Islands		
69	Northern Marianas		
70	Palau		
72	Puerto Rico		
78	Virgin Islands


- OBEREG - Bureau of Economic Analysis (BEA) Regions
BEA Regions are a set of Geographic Areas that are aggregations of the states. The regional classifications, which were developed in the mid-1950s, are based on the homogeneity of the states in terms of economic characteristics, such as the industrial composition of the labor force, and in terms of demographic, social, and cultural characteristics. BEA groups all 50 states and the District of Columbia into eight distinct regions for purposes of data collecting and analyses. IPEDS/HEGIS assigns U.S. service schools as their own region.

0 - US Service schools
1 - New England CT ME MA NH RI VT
2 - Mid East DE DC MD NJ NY PA
3 - Great Lakes IL IN MI OH WI
4 - Plains IA KS MN MO NE ND SD
5 - Southeast AL AR FL GA KY LA MS NC SC TN VA WV
6 - Southwest AZ NM OK TX
7 - Rocky Mountains CO ID MT UT WY
8 - Far West AK CA HI NV OR WA
9 - Outlying areas AS FM GU MH MP PR PW VI
-3 - Not available

- CHFNM - Name of chief administrator of institution

-1 - Not reported
-2 - Not applicable
-9 - not active

- CHFTITLE
Title of chief administrator of institution
-1 - Not reported
-2 - Not applicable
-9 - not active

- GENTELE 
General information telephone number
-1 - Not reported
-2 - Not applicable
-9 - not active


- EIN - Employer Identification Number
The number assigned to an institution by the Internal Revenue Service for tax purposes.

- DUNS 
The Data Universal Numbering System (DUNS), is a proprietary system developed and regulated by Dun & Bradstreet (D&B) that assigns a unique numeric identifier, referred to as a "DUNS number" to a single business entity. It was introduced in 1963 to support D&B's credit reporting practice. The DUNS number is a 9-digit numeric. Some institutions have multiple DUNS numbers. Multiple DUNS numbers are separated by the vertical bar.


- OPEID  
Identification number used by the U.S. Department of Education's Office of Postsecondary Education (OPE) to identify schools that have Program Participation Agreements (PPA) so that its students are eligible to participate in Federal Student Financial Assistance programs under Title IV regulations. This is a 6-digit number followed by a 2-digit suffix used to identify branches, additional locations, and other entities that are part of the eligible institution.


- OPEFLAG

Code indicating the institution's degree of eligibility for Title IV aid.

1	Participates in Title IV federal financial aid programs		
2	Branch campus of a main campus that participates in Title IV		
3	Deferment only - limited participation		
4	New participants (became eligible during winter collection)		
8	New participants (became eligible during spring collection)		
5	Not currently participating in Title IV, has an OPE ID number		
6	Not currently participating in Title IV, does not have OPE ID number		
7	Stopped participating during the survey year		


- WEBADDR
Institution's internet website address


- ADMINURL
Admissions office web (URL) address


- FAIDURL
Financial aid office web (URL) address

- APPLURL
Web address (URL) on online applciation


- NPRICURL
Net Price Calculator Web Address for institutions that enroll full-time, first-time undergraduate students

- VETURL
Veterans and Military Servicemembers tuition policies web address

Does your institution post tuition policies specifically related to Veterans and Military Servicemembers on a website? Provide URL

- ATHURL
Student-Right-to-Know student athlete graduation rate website URL


- DISAURL
Disability Services Web Address


- SECTOR
Sector of institution

Value	Label	Frequency	Percentage
0	Administrative Unit		
1	Public, 4-year or above		
2	Private not-for-profit, 4-year or above		
3	Private for-profit, 4-year or above		
4	Public, 2-year		
5	Private not-for-profit, 2-year		
6	Private for-profit, 2-year		
7	Public, less-than 2-year		
8	Private not-for-profit, less-than 2-year		
9	Private for-profit, less-than 2-year		
99	Sector unknown (not active)		

One of nine institutional categories resulting from dividing the universe according to control and level. Control categories are public, private not-for-profit, and private for-profit. Level categories are 4-year and higher (4 year), 2-but-less-than 4-year (2 year), and less than 2-year. For example: public, 4-year institutions.

Control - A classification of whether an institution is operated by publicly elected or appointed officials (public control) or by privately elected or appointed officials and derives its major source of funds from private sources (private control).

Level - A classification of whether an institution’s programs are 4-year or higher (4 year), 2-but-less-than 4-year (2 year), or less than 2-year.


- ICLEVEL
A classification of whether an institution’s programs are 4-year or higher (4 year), 2-but-less-than 4-year (2 year), or less than 2-year.

1	Four or more years		
2	At least 2 but less than 4 years		
3	Less than 2 years (below associate)		
-3	{Not available}		



- CONTROL

A classification of whether an institution is operated by publicly elected or appointed officials or by privately elected or appointed officials and derives its major source of funds from private sources.

Public institution - An educational institution whose programs and activities are operated by publicly elected or appointed school officials and which is supported primarily by public funds.

Private not-for-profit institution - A private institution in which the individual(s) or agency in control receives no compensation, other than wages, rent, or other expenses for the assumption of risk. These include both independent not-for-profit schools and those affiliated with a religious organization.

Private for-profit institution - A private institution in which the individual(s) or agency in control receives compensation other than wages, rent, or other expenses for the assumption of risk.

1	Public		
2	Private not-for-profit		
3	Private for-profit		
-3	{Not available}




- HLOFFER
Highest level of offering

0 Other
1	Award of less than one academic year		
2	At least 1, but less than 2 academic yrs		
3	Associate's degree		
4	At least 2, but less than 4 academic yrs		
5	Bachelor's degree		
6	Postbaccalaureate certificate		
7	Master's degree		
8	Post-master's certificate		
9	Doctor's degree		
-2	Not applicable, first-professional only		
-3	{Not available}


- UGOFFER

A code indicating whether the institution offers undergraduate degrees or certificates.
	
Undergraduate offering
Note: The listed values are based on the whole IPEDS universe and don't represent your group.
Value	Label	Frequency	Percentage
1	Undergraduate degree or certificate offering		
2	No undergraduate offering		
-3	{Not available}


- GROFFER
A code indicating whether the institution offers graduate degrees or certificates.

Graduate degrees or certificates include master^s and doctor's degrees and postbaccalaureate and post-master's certificates

Graduate offering
Note: The listed values are based on the whole IPEDS universe and don't represent your group.
Value	Label	Frequency	Percentage
1	Graduate degree or certificate offering		
2	No graduate offering		
-3	{Not available}



- HDEGOFR1

A code for highest degree offered at the institution - This variable was revised in 2009 because the first professional category is no longer an award level and the Doctoral award level category has expanded to 3 categories, Doctor's degree - Research/scholarships, Doctor's degree - professional practice and Doctor's degree-other.

This variable is derived from the level of offering variables (LEVEL1-LEVEL8 LEVEL12 and LEVEL17-LEVEL19) from the Institutional Characteristics component as follows:

IF LEVEL17=1 and LEVEL18=1 then HDEGOFR1=11
else if LEVEL17=1 and LEVEL18 ne 1 then HDEGOFR1=12
else if LEVEL17 ne 1 and LEVEL18=1 then HDEGOFR1=13
else if LEVEL17 ne 1 and LEVEL18 ne 1 and LEVEL19=1 then HDEGOFR1=14
else if LEVEL7=1 HDEGOFR1=20
else if LEVEL5=1 HDEGOFR1=30
else if LEVEL3=1 HDEGOFR1=40
else if LEVEL1=1 or LEVEL2=1 or LEVEL4=1 or LEVEL6=1 or LEVEL8=1
or LEVEL12=1 then HDEGOFR1=0
else HDEGOFR1=-3;
For more information see descriptions of LEVEL1-LEVEL12

11	Doctor's degree - research/scholarship and professional practice		
12	Doctor's degree - research/scholarship		
13	Doctor's degree - professional practice		
14	Doctor's degree - other		
20	Master's degree		
30	Bachelor's degree		
40	Associate's degree		
0	Non-degree granting		
-3	{Not available}



- DEGGRANT
A code indicating the degree-granting status of the institution.

Degree-granting institutions offer an associate's, bachelor's, master's, doctor's or a first-professional degree.

Nondegree-granting offers certificates or other formal awards.

	
Degree-granting status
Note: The listed values are based on the whole IPEDS universe and don't represent your group.
Value	Label	Frequency	Percentage
1	Degree-granting		
2	Nondegree-granting, primarily postsecondary		
-3	{Not available}



- HBCU
A code to indicate whether the institution is one of the Historically Black College or University (HBCU) institutions.

Historically Black Colleges and Universities (HBCU) - The Higher Education Act of 1965, as amended, defines an HBCU as: "...any historically black college or university that was established prior to 1964, whose principal mission was, and is, the education of black Americans, and that is accredited by a nationally recognized accrediting agency or association determined by the Secretary [of Education] to be a reliable authority as to the quality of training offered or is, according to such an agency or association, making reasonable progress toward accreditation." Federal regulations (20 USC 1061 (2)) allow for certain exceptions to the founding date.

1	Yes		
2	No


- HOSPITAL
A code to indicate whether the institution has hospital.
1	Yes		
2	No		
-1	Not reported		
-2	Not applicable	



- MEDICAL 
A code to indicate whether the institution grants a medical degree.

This variable is derived from the 2012-13 completions component of the IPEDS survey.

Medical degrees include:
Medicine (M.D.) (51.1201),
Dentistry (D.D.S., D.M.D.) (51.0401)
Osteopathic Medicine (D.O.) (51.1901)
Veterinary Medicine (D.V.M.) (51.2401)

1	Yes		
2	No		
-1	Not reported		
-2	Not applicable	




- TRIBAL
A code to indicate whether the institution is one of the Tribal Colleges and Universities. These institutions, with few exceptions, are tribally controlled and located on reservations. They are all members of the American Indian Higher Education Consortium.

1	Yes		
2	No




- LOCALE
Locale codes identify the geographic status of a school on an urban continuum ranging from “large city” to “rural.” They are based on a school’s physical address. The urban-centric locale codes introduced in this file are assigned through a methodology developed by the U.S. Census Bureau’s Population Division in 2005.

11	City: Large		
12	City: Midsize		
13	City: Small		
21	Suburb: Large		
22	Suburb: Midsize		
23	Suburb: Small		
31	Town: Fringe		
32	Town: Distant		
33	Town: Remote		
41	Rural: Fringe		
42	Rural: Distant		
43	Rural: Remote		
-3	{Not available}



- OPENPUBL
A code to indicate whether the institution is open for admission to the general public.

1	Institution is open to the public		
0	Institution is not open to the public	



- ACT
A code to indicate the status of institution's participation in IPEDS.

A	Active - institution active		
N	New (active) - added during the current year		
R	Restore (active) - restored to the current universe		
M	Closed in current year (active has data)		
C	Combined with other institution		
D	Delete out of business		
I	Inactive due to hurricane related problems		
O	Out-of-scope of IPEDS - not postsecondary		
P	Potential new/add institution		
Q	Potential restore institution		
W	Potential add not within scope of IPEDS		
X	Potential restore not within scope of IPEDS		
G	Perfect child campus - data reported with main campus, but is listed on college navigator




- NEWID



- DEATHYR
Year institution was deleted from IPEDS.
2021	2021		
-2	Not applicable




"CLOSEDAT
Date institution closed - format is mmddyyyy



CYACTIVE
A code to indicate whether the institution is active in current year. This code is derived from the more detailed ACT (status of institution) variable.

Active school CYACTIVE=1 includes:
ACT=A an open (active) institution
ACT=N an institution that opened in the current year
ACT=M an institution that closed in the current year, but will have data from the previous academic year for example Finance or Completions data for academic year 2016-17
ACT=R an institution that was previously not active and was restored in the current year

CYACTIVE=2 includes potential adds and restores (ACT=P,R)

CYACTIVE=3 includes all other inactive institutions
ACT=C combined with another active institution
ACT=D delete institution is out of business
ACT=O institution is out-of-scope
ACT=W potential add that was not added
ACT=X potential restore that was not restored

1	Yes		
2	No, potential add or restore		
3	No, closed, combined, or out-of-scope	




- POSTSEC
A derived indicator to easily select institutions who are (1) currently in the IPEDS universe, (2) open to the public, (3) participant in Title IV program, and (4) primarily postsecondary institution.

1	Title IV postsecondary institution		
2	Non-Title IV postsecondary institution		
3	Title IV NOT primarily postsecondary institution		
4	Non-Title IV NOT primarily postsecondary institution		
5	Title IV postsecondary institution that is NOT open to the public		
6	Non-Title IV postsecondary institution that is NOT open to the public		
9	Institution is not active in current universe		






"PSEFLAG
Postsecondary institution indicator
Identifies an institution whose primary purpose is to provide postsecondary education,is open to the general public and is currently an active (operating) institution.

1	Active postsecondary institution		
2	Not primarily postsecondary or open to public		
3	Not active



- PSET4FLG
A derived indicator to easily select institutions who are (1) currently in the IPEDS universe, (2) open to the public, (3) participant in Title IV program, and (4) primarily postsecondary institution.

1	Title IV postsecondary institution		
2	Non-Title IV postsecondary institution		
3	Title IV NOT primarily postsecondary institution		
4	Non-Title IV NOT primarily postsecondary institution		
5	Title IV postsecondary institution that is NOT open to the public		
6	Non-Title IV postsecondary institution that is NOT open to the public		
9	Institution is not active in current universe




- RPTMTH
Reporting method identifies how institutions report student charges (full academic year or by program) and the type of cohort (fall or full-year) used to report graduation rate, retention rate and student financial aid data.
1	Student charges for full academic year and fall GR/SFA/retention rate cohort		
3	Student charges for full academic year and full-year GR/SFA/retention rate cohort		
2	Student charges by program and full-year GR/SFA/retention rate cohort		
-1	Not reported		
-2	Not applicable




- INSTCAT
Institutional category was derived using the level of offerings reported on the Institutional Characteristics (IC) component and the number and level of awards that were reported on the Completions (C) component.

1	Degree-granting, graduate with no undergraduate degrees		
2	Degree-granting, primarily baccalaureate or above		
3	Degree-granting, not primarily baccalaureate or above		
4	Degree-granting, associate's and certificates		
5	Nondegree-granting, above the baccalaureate		
6	Nondegree-granting, sub-baccalaureate		
-1	Not reported		
-2	Not applicable




- C18BASIC
The Basic Classification is an update of the traditional classification framework developed by the Carnegie Commission on Higher Education in 1970 to support its research program. The Basic Classification was originally published for public use in 1973, and subsequently updated in 1976, 1987, 1994, 2000, 2005, 2010, 2015 and 2018. In the 2018 update, the Doctoral Universities have been reshaped to better accommodate “Doctor's degree – professional practice” within our methodology. Please see the Basic Classification Methodology for details regarding how this classification is calculated.

Doctoral universities - Includes institutions that awarded at least 20 research/scholarship doctoral degrees during the update year and also institutions with below 20 research/scholarship doctoral degrees that awarded at least 30 professional practice doctoral degrees in at least 2 programs. Excludes Special Focus Institutions and Tribal Colleges.

The first two categories include only institutions that awarded at least 20 research/scholarship doctoral degrees and had at least $5 million in total research expenditures (as reported through the National Science Foundation (NSF) Higher Education Research & Development Survey (HERD)).

R1: Doctoral Universities – Very high research activity
R2: Doctoral Universities – High research activity
D/PU: Doctoral/Professional Universities

Master's Colleges and Universities - Generally includes institutions that awarded at least 50 master's degrees and fewer than 20 doctoral degrees during the update year (with occasional exceptions – see Methodology). Excludes Special Focus Institutions and Tribal Colleges.

Baccalaureate Colleges - Includes institutions where baccalaureate or higher degrees represent at least 50 percent of all degrees but where fewer than 50 master's degrees or 20 doctoral degrees were awarded during the update year. (Some institutions above the master's degree threshold are also included; see Methodology.) Excludes Special Focus Institutions and Tribal Colleges.

Baccalaureate/Associate's Colleges - Includes four-year colleges (by virtue of having at least one baccalaureate degree program) that conferred more than 50 percent of degrees at the associate's level. Excludes Special Focus Institutions and Tribal Colleges.

Associate's Colleges - Institutions at which the highest level degree awarded is an associate's degree. The institutions are sorted into nine categories based on the intersection of two factors: disciplinary focus (transfer, career & technical or mixed) and dominant student type (traditional, nontraditional or mixed). Excludes Special Focus Institutions and Tribal Colleges.

Special Focus Institutions- Institutions where a high concentration of degrees is in a single field or set of related fields. Excludes Tribal Colleges.

Tribal Colleges - Colleges and universities that are members of the American Indian Higher Education Consortium, as identified in IPEDS Institutional Characteristics.

Classifications are time-specific snapshots of institutional attributes and behavior based on 2016-17 data.

The methodology can be found at http://carnegieclassifications.iu.edu/definitions.php.


1	Associate's Colleges: High Transfer-High Traditional		
2	Associate's Colleges: High Transfer-Mixed Traditional/Nontraditional		
3	Associate's Colleges: High Transfer-High Nontraditional		
4	Associate's Colleges: Mixed Transfer/Career & Technical-High Traditional		
5	Associate's Colleges: Mixed Transfer/Career & Technical-Mixed Traditional/Nontraditional		
6	Associate's Colleges: Mixed Transfer/Career & Technical-High Nontraditional		
7	Associate's Colleges: High Career & Technical-High Traditional		
8	Associate's Colleges: High Career & Technical-Mixed Traditional/Nontraditional		
9	Associate's Colleges: High Career & Technical-High Nontraditional		
10	Special Focus Two-Year: Health Professions		
11	Special Focus Two-Year: Technical Professions		
12	Special Focus Two-Year: Arts & Design		
13	Special Focus Two-Year: Other Fields		
14	Baccalaureate/Associate's Colleges: Associate's Dominant		
15	Doctoral Universities: Very High Research Activity		
16	Doctoral Universities: High Research Activity		
17	Doctoral/Professional Universities		
18	Master's Colleges & Universities: Larger Programs		
19	Master's Colleges & Universities: Medium Programs		
20	Master's Colleges & Universities: Small Programs		
21	Baccalaureate Colleges: Arts & Sciences Focus		
22	Baccalaureate Colleges: Diverse Fields		
23	Baccalaureate/Associate's Colleges: Mixed Baccalaureate/Associate's		
24	Special Focus Four-Year: Faith-Related Institutions		
25	Special Focus Four-Year: Medical Schools & Centers		
26	Special Focus Four-Year: Other Health Professions Schools		
27	Special Focus Four-Year: Engineering Schools		
28	Special Focus Four-Year: Other Technology-Related Schools		
29	Special Focus Four-Year: Business & Management Schools		
30	Special Focus Four-Year: Arts, Music & Design Schools		
31	Special Focus Four-Year: Law Schools		
32	Special Focus Four-Year: Other Special Focus Institutions		
33	Tribal Colleges		
0	(Not classified)		
-2	Not applicable, not in Carnegie universe (not accredited or nondegree-granting)






- C18IPUG






"C18IPGRD" 


"C18UGPRF"


- "C18ENPRF"

"C18SZSET"

"C15BASIC

1	Associate's Colleges: High Transfer-High Traditional		
2	Associate's Colleges: High Transfer-Mixed Traditional/Nontraditional		
3	Associate's Colleges: High Transfer-High Nontraditional		
4	Associate's Colleges: Mixed Transfer/Career & Technical-High Traditional		
5	Associate's Colleges: Mixed Transfer/Career & Technical-Mixed Traditional/Nontraditional		
6	Associate's Colleges: Mixed Transfer/Career & Technical-High Nontraditional		
7	Associate's Colleges: High Career & Technical-High Traditional		
8	Associate's Colleges: High Career & Technical-Mixed Traditional/Nontraditional		
9	Associate's Colleges: High Career & Technical-High Nontraditional		
10	Special Focus Two-Year: Health Professions		
11	Special Focus Two-Year: Technical Professions		
12	Special Focus Two-Year: Arts & Design		
13	Special Focus Two-Year: Other Fields		
14	Baccalaureate/Associate's Colleges: Associate's Dominant		
15	Doctoral Universities: Highest Research Activity		
16	Doctoral Universities: Higher Research Activity		
17	Doctoral Universities: Moderate Research Activity		
18	Master's Colleges & Universities: Larger Programs		
19	Master's Colleges & Universities: Medium Programs		
20	Master's Colleges & Universities: Small Programs		
21	Baccalaureate Colleges: Arts & Sciences Focus		
22	Baccalaureate Colleges: Diverse Fields		
23	Baccalaureate/Associate's Colleges: Mixed Baccalaureate/Associate's		
24	Special Focus Four-Year: Faith-Related Institutions		
25	Special Focus Four-Year: Medical Schools & Centers		
26	Special Focus Four-Year: Other Health Professions Schools		
27	Special Focus Four-Year: Engineering Schools		
28	Special Focus Four-Year: Other Technology-Related Schools		
29	Special Focus Four-Year: Business & Management Schools		
30	Special Focus Four-Year: Arts, Music & Design Schools		
31	Special Focus Four-Year: Law Schools		
32	Special Focus Four-Year: Other Special Focus Institutions		
33	Tribal Colleges		
0	(Not classified)		
-2	Not applicable, not in Carnegie universe (not accredited or nondegree-granting)		
Variable Description
The Basic Classification is an update of the traditional classification framework developed by the Carnegie Commission on Higher Education in 1970 to support its research program. The Basic Classification was published for use in 1973, and subsequently updated in 1976, 1987, 1994, 2000, 2005 and 2010 and 2015. The 2005 edition involved some significant changes from the previous releases by dividing Associate's colleges into subcategories and using a multi-measure research index to classify doctorate-granting institutions. The 2010 update retained the same classification structure as the 2005 edition. In the 2015 update, the Associate's Categories were substantially redefined and the categories of the Research Doctoral Universities changed (but not the calculation methodology). Please see the Basic Classification Methodology for details regarding how this classification is calculated.

Note: The "shorthand" labels for the Doctoral Universities and Master's Colleges and Universities were restored in the 2015 update to numeric sequences (R1, R2, R3, and M1, M2, M3) to denote that each one is based on differences in quantitative levels. For doctoral universities, the levels are based on a research activity index and for master's colleges and universities it is based on number of degrees conferred.

Doctoral Universities - Includes institutions that awarded at least 20 research/scholarship doctoral degrees during the update year (this does not include professional practice doctoral-level degrees, such as the JD, MD, PharmD, DPT, etc.). Excludes Special Focus Institutions and Tribal Colleges.

Master's Colleges and Universities - Generally includes institutions that awarded at least 50 master's degrees and fewer than 20 doctoral degrees during the update year (with occasional exceptions – see Methodology). Excludes Special Focus Institutions and Tribal Colleges.

Baccalaureate Colleges - Includes institutions where baccalaureate or higher degrees represent at least 50 percent of all degrees but where fewer than 50 master's degrees or 20 doctoral degrees were awarded during the update year. (Some institutions above the master's degree threshold are also included; see Methodology.) Excludes Special Focus Institutions and Tribal Colleges.

Baccalaureate/Associate's Colleges - Includes four-year colleges (by virtue of having at least one baccalaureate degree program) with fewer than 50 percent of degrees awarded at the associate's level. Excludes Special Focus Institutions and Tribal Colleges.

Associate's Colleges - Institutions at which the highest level degree awarded is an associate's degree. The institutions are sorted into nine categories based on the intersection of two factors: disciplinary focus (transfer, career & technical or mixed) and dominant student type (traditional, nontraditional or mixed). Excludes Special Focus Institutions and Tribal Colleges.

Special Focus Institutions- Institutions where a high concentration of degrees is in a single field or set of related fields. Excludes Tribal Colleges.

Tribal Colleges - Colleges and universities that are members of the American Indian Higher Education Consortium, as identified in IPEDS Institutional Characteristics.

Classifications are time-specific snapshots of institutional attributes and behavior based on 2013-14 data.

The methodology can be found at http://carnegieclassifications.iu.edu/methodology/basic.php.



"CCBASIC"  "CARNEGIE" "LANDGRNT" "INSTSIZE" "F1SYSTYP" "F1SYSNAM" "F1SYSCOD" "CBSA"     "CBSATYPE" "CSA"     
[66] "NECTA"    "COUNTYCD" "COUNTYNM" "CNGDSTCD" "LONGITUD" "LATITUDE" "DFRCGID"  "DFRCUSCG"
