---
title: "Array Dimensions"
author: "Kehinde Fagbamigbe"
date: "2023-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
```


- Y[i,2,j,k,t]
- #i, no of records
- #j/p Majors
- #k/q: Race
- #t/T Year



- Y[i,2,j,k,t]
- #i, no of records
- #j/p Majors
- #k/q: Race
- #t/T Year

998   2   4   5  13


### Demo
```{r}
dim_names <- list(c("Inst1", "Inst2", "Inst3"), c("M","F"), c("Maths", "Phy", "Bio"), c("Black", "Hispanic", "White", "Latino"), c("yr1","yr2","yr3","yr4","yr5"))
my_5d_array <- array(1:450, c(3, 2, 3, 4, 5), dimnames = dim_names)
my_5d_array
```
- Y[i,2,j,k,t]
- #i, no of records
- #j/p Majors
- #k/q: Race
- #t/T Year

998   2   4   5  13

## this is the data for year 2!!!

```{r}
year2 <- my_5d_array[ , , , , 2]
dim(year2)
year2
```

### this is the data for just the female gender!
dim_names <- list(c("Inst1", "Inst2", "Inst3"), c("M","F"), c("Maths", "Phy", "Bio"), c("Black", "Hispanic", "White", "Latino"), c("yr1","yr2","yr3","yr4","yr5"))

```{r}
females <- my_5d_array[, 2, , , ]
dim(females)
females
```
```{r}
percent_missing <- function(array) {
  total_elements <- prod(dim(array))
  missing_elements <- sum(is.na(array))
  return(missing_elements/total_elements * 100)
}

```


```{r}
percent_missing(my_5d_array)
```

```{r}
dim(females)
```
- Y[i,2,j,k,t]
- #i, no of records
- #j/p Majors
- #k/q: Race
- #t/T Year

998   2   4   5  13


```{r}
female_per_year <- apply(females, 4, sum) #sums across the individual years
female_per_year
```
```{r}
dim(female_per_year)
```

```{r}
class(female_per_year)
```

```{r}
colnames(female_per_year)
```

```{r}
female_sum_apply <- apply(females, 3, sum)
female_sum_apply
```



```{r}
female_race_per_year <- apply(females, 3:4, sum) #no of race per year for females
female_race_per_year
```

```{r}
female_per_institution_per_year <- apply(females, c(1,4), sum) #no of females in institution per year 
female_per_institution_per_year
```

```{r}
dim(female_per_institution_per_year)
```


```{r}
class(female_per_institution_per_year)
```


```{r}
#image(female_per_institution_per_year)
```


```{r}
#matplot(female_per_institution_per_year)
```

```{r}
female_per_institution_per_year
```

```{r}
ncol(female_per_institution_per_year)
```

```{r}
vector <- as.vector(female_per_institution_per_year)
vector
```

```{r}
# Convert the 2D array into a data frame
# df <- data.frame(x = 1:nrow(female_sum_apply), y = 1:ncol(female_sum_apply), z = as.vector(female_sum_apply))
# df
```


### Converting matrix or array to a dataframe
```{r}
df <- as.data.frame(female_per_institution_per_year)
df
```

### Obtaining the row names
```{r}
rownames(df)
```


```{r}
colnames(df)
```


### Add the index as a column
```{r}
df_long <- df %>%
  mutate(Inst = row.names(df))
df_long
#rownames(df_long) <- NULL
df_long
```

# Convert the data frame to a long format for plotting
```{r}
df_long <- reshape2::melt(df_long, id.vars = "Inst")
df_long
```



```{r}
# Plot the grouped bar chart
ggplot(df_long, aes(x = Inst, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") + 
  #geom_text(aes(label = value)) +
  xlab("Institution") +
  ylab("No of Females") +
  ggtitle("Comparison of Females in Institutions by Year") +
  theme(legend.title = element_blank())

```

### Add the index as a column
```{r}
df_long <- df %>%
  mutate(Inst = row.names(df))
df_long
rownames(df_long) <- NULL
df_long
```
```{r}
df_long <- reshape2::melt(df_long, id.vars = "Inst")
df_long
```

```{r}
# Plot the grouped bar chart
ggplot(df_long, aes(x = Inst, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") + 
  #geom_text(aes(label = value)) +
  xlab("Institution") +
  ylab("No of Females") +
  ggtitle("Comparison of Females in Institutions by Year") +
  theme(legend.title = element_blank())

```


```{r}
# Plot the grouped bar chart
ggplot(df_long, aes(x = Inst, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Institution") +
  ylab("No of Females") +
  ggtitle("Comparison of Females in Institutions by Year") +
  theme(legend.title = element_blank())
```

```{r}
race <- apply(females, 3, sum)
race
```

```{r}
class(race)
```


### Converting to a dataframe
```{r}
race_df <- as.data.frame(race)
race_df
```


```{r}
race_df <- race_df %>%
  mutate(Races = row.names(race_df))
race_df
rownames(race_df) <- NULL
race_df
```



```{r}
# Plot the bar chart
ggplot(race_df, aes(x = Races, y = race)) +
  geom_bar(stat = "identity", fill = "purple") +
  xlab("Race") +
  ylab("No of Females") +
  ggtitle("Race Comparison for Females") +
  theme(legend.title = element_blank())

```

def find_missing_values(array):
    missing_values = []
    for i in range(array.shape[0]):
        for j in range(array.shape[1]):
            for k in range(array.shape[2]):
                for l in range(array.shape[3]):
                    for m in range(array.shape[4]):
                        if np.isnan(array[i][j][k][l][m]):
                            missing_values.append((i, j, k, l, m))
    return missing_values
    
    
    

```{r}
dim(my_5d_array)
```


```{r}
find_missing_values <- function(my_5d_array) {
  dims <- dim(my_5d_array)
  missing_values <- which(is.na(my_5d_array), arr.ind=TRUE)
  return(missing_values)
}

```


```{r}
find_missing_values(my_5d_array)
```

```{r}
dim_names <- list(c("Inst1", "Inst2", "Inst3"), c("M","F"), c("1", "2", "3"), c("yr1","yr2"))
my_5d_array <- array(1:27, c(3, 2, 3, 2), dimnames = dim_names)
my_5d_array
```

```{r}
dim_names <- list(c("a", "b", "c"), c("x", "y", "z"), c("1", "2", "3"))
my_3d_array <- array(1:27, c(3, 3, 3), dimnames = dim_names)
my_3d_array
```




```{r}
my_3d_array[, , ]
```

```{r}
my_3d_array[, , 1:2]
```


```{r}
my_3d_array[, ,1]
```


```{r}
my_3d_array[, 1,1]
```


```{r}
my_3d_array[, ,1][,1:2]
```

```{r}
my_3d_array[, ,1][,c("x", "z")]
```


```{r}
dim_names <- list(c("a", "b"), c("w","x", "y", "z"), c("1", "2", "3"))
my_array <- array(1:27, c(2, 4, 3), dimnames = dim_names)
my_array
```


```{r}
dim_names <- list(c("a", "b"), c("w","x", "y", "z"), c("1", "2", "3"), c("p","q"))
my_5d_array <- array(1:27, c(2, 4, 3, 2), dimnames = dim_names)
my_5d_array
```